Tidy data structure to support exploration and modeling of temporal-context data

### 2018-06-02 main.Rmd
mapped out the literature in data pipeline

@wickham2014tidy coined the term "tidy data" and formalized the processing from messy data to "tidy data". These principles attempt to standardize the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. Based on the systematic structuring principles, a grammar of data manipulation. [R]

@@

@@

The "key" identifies units or subjects that are observed over time in a data table, which are typically known in advance by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter; on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other. Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of missing combinations.

@@

Literature about data pipeline.

A pipeline exhibits a hierarchy of data operations: (1) atomic (1-dimensional) vectors (`mean(variable)`) --> (2) (2-dimensional) data table (`summarise()`) --> repeat step (1) and (2) to form a chain. Reversely, a data pipeline is decomposed into rectangular blocks, and then into atomic strips. (UNIX pipe)

@@

* **row-wise**: `filter()`, `slice()`, `arrange()`, `fill_na()`
* **column-wise**: `mutate()`, `select()`, `summarise()`
* **group-wise**: `index_by()`, `group_by()`
* **rolling window**: `slide()`, `tile()`, `stretch()`
* **time-wise**: `lead()`, `lag()`, `difference()`

>>>>

@wickham2014tidy coined the term "tidy data" and formalized the processing from messy data to "tidy data". These principles attempt to standardize the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. Based on the systematic structuring principles, a grammar of data manipulation. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools. [R]

@@

[A toy example is needed here and will referenced for index and key]

@@

The "key" identifies units or subjects that are observed over time in a data table, which are typically known in advance by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter; on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other. Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.

@@

## Literature review

* A collection of intuitive and consistent programming interfaces empowers users to focus on the primary goal of data analysis and modelling problems, and frees users from computational implementation details.
* [@Unwin99guiand] since programming interfaces are precise and repeatable, they are preferable when we can describe exactly what we want, but a GUI is better when: “Searching for information and interesting structures without fully specified questions.” [R]
* Unix philosophy [@unix; @unix-philosophy]: each do one simple thing, do it well, and most importantly, work well with each other.
* "No matter how complex and polished the individual operations are, it is often the quality of the glue that most directly determines the power of the system" [@eopl]

* Pipelines in interactive settings [@viewing-pipeline; @plumbing; @xie2014reactive]: Model-View-Controller paradigm.

## Time-based pipeline analysis

A pipeline exhibits a hierarchy of data operations: (1) atomic (1-dimensional) vectors (`mean(variable)`) --> (2) (2-dimensional) data table (`summarise()`) --> repeat step (1) and (2) to form a chain. Reversely, a data pipeline is decomposed into rectangular blocks, and then into atomic strips.

@@

* **row-wise**: `filter()`, `slice()`, `arrange()`, `fill_na()`
* **column-wise**: `mutate()`, `select()`, `summarise()`
* **group-wise**: `index_by()`, `group_by()`
* **time-wise**: `lead()`, `lag()`, `difference()`
* **rolling window**: `slide()`, `tile()`, `stretch()`
